---
title: "Interim Data Analysis Plan"
author: |
    | Analyst: Max McGrath
    | Investigator: Dr. Marty Otañez
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
    - \usepackage{fixmath}
    - \usepackage{amsmath}
output: pdf_document
indent: false
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
library(kableExtra)
setwd("..")
source("Code/4_summaryTables.R")
```

## Introduction

We received responses to 102 survey questions from
213 individuals who currently work in the Colorado cannabis industry or who were
employed in the Colorado cannabis industry in the past 12 months. 
Most relevant to this analysis, respondents were asked whether they support
unionization within the cannabis industry for three purposes: creation of safer
workplaces, negotiation of better wages, and negotiation of better benefits.
Further, respondent's ability to recognize existing federal labor protections was also
assessed, and they were asked whether they have concerns regarding their exposure
to three types of workplace hazards: biological, chemical, and physical. Lastly,
they were asked their age, gender, race, and hourly compensation.

\qquad This study aims to investigate the relationships between cannabis employees' 
support and knowledge of labor rights and their demographics, wages, and
concerns over exposure to occupational hazards by testing the following four
hypotheses:

1. Cannabis workers who express concern over exposure to occupational hazards are
more likely to express support for labor rights.
2. Female cannabis workers are more likely to support workers’ rights than their
male counterparts.
3. Non-white cannabis workers are more likely to support workers’ rights than 
their white counterparts.
4. Higher paid cannabis employees show more knowledge about workers’ rights than
lower paid employees.

## Methods

### Data Preparation

First, incomplete observations were removed from the data set. These incomplete
observations came from the questions about gender and race, as responses
where individuals chose "prefer not to say" for either question 
were marked as missing. All other relevant question had no missing 
responses. Table 1 describes the missingness in the observed data set.

\qquad It is also notable that the question asking survey takers' gender identity
included a 
"non-binary/third gender" option. So, for the purposes of this analysis, the gender 
category was reduced to a binary category in which respondents were classified 
as those who identify as female and those who do not. Similarly, the question 
that asked respondents to describe their race
included seven options, and that data was reduced to a binary
category classifying respondents as those who describe themselves as a race 
other than white (non-white) and those who describe themselves as white. Age
and hourly compensation (wage) are both categorical variables with 6 and 7
possible responses, respectively. Lists of possible responses is included as a
part of Tables 2 and 3. 

\qquad Measurements of workers' concerns over occupational hazards stem from 
their answers to the following three multiple response questions: 

* What __biological__ hazards exist in your cannabis workplace that concern
you?
* What __chemical__ hazards exist in your cannabis workplace that concern you?
* What __physical__ hazards exist in your cannabis workplace that concern you? 

Each of these questions allowed users to select multiple responses from a list
of relevant types of hazards contained within each category. For the purposes of
this analysis, each respondent
was given a binary assignment for each category of hazard which indicates 
whether they expressed concern for any hazard in that category. A complete
description of each category of hazard is provided in Table 4. 

\qquad Cannabis workers' support for unionism was measured based on their 
responses to the three questions:

* Do you believe that cannabis workers need to be part of a labor union
union to __create safer workplaces__?
* Do you believe that cannabis workers need to be part of a labor union
union to __negotiate better wages__?
* Do you believe that cannabis workers need to be part of a labor union
union to __negotiate better benefits__?

Each of these questions had two possible responses, "yes" and "no". Each
question's set of responses will be analyzed separately so that
differing motivations for support of unionism may be elucidated by this
analysis. A complete description of respondents' answers to these questions is
provided in Table 2, with counts stratified by values for each the above
covariates.

\qquad Lastly, respondents' knowledge of workers rights was assessed
by their ability to recognize four existing federal protections. The
individuals surveyed were asked:

* Does federal law protect workers under the worker protection standard 
from retaliation from bringing forward safety concerns?
* Does federal law allow cannabis workers to form a labor union?
* Does federal labor law prohibit cannabis employers from taking any 
adverse employment action against you for engaging in concerted activity
regarding wages, benefits and other terms and conditions of 
employment?
* Do federal labor laws protect you in engaging in concerted activity even 
if you are not in a labor union?

Each question had response options of "yes", "no", and "I don't know". The
respondents were provided an example of concerted activity which says, 
"complaining about safety, having discussion with co-workers about working 
conditions."
Noting that the correct response to each question is "yes", each respondent was
assigned a score from 0 to 4, with one point being awarded for each correct
response. Table 3 provides the sample mean (and standard error) for assigned
scores stratified by each covariate of interest.

### Analysis

To evaluate Hypotheses 1-3, three logistic regression models, one for each type
of unionism support,
will undergo model selection. Then, we will use each
model's final coefficients and their corresponding estimates and
statistical significance to determine whether the data falsifies or lends 
evidence to each of the above hypotheses.

\qquad A logistic regression model is specified as

$$
P(X)=\frac{e^{\beta_0+\beta_1X_1+\beta_2X_2+\cdots+\beta_pX_p}}{1+e^{\beta_0+\beta_1X_1+\beta_2X_2+\cdots+\beta_pX_p}}
$$

where $P(X)=P(Y=1|X)$ represents the probability that an individual belongs to 
a certain outcome class given their observed covariate values. In this analysis,
we will use binary outcomes ("yes" = 1, "no" = 0). 
The covariates
in a logistic regression model may be continuous, ordinal, or binary, but we 
will only
be using binary and ordinal covariates. The above
equation is typically rewritten as

$$
\log(\frac{P(X)}{1-P(X)})=\beta_0+\beta_1X_1+\cdots+\beta_pX_p
$$

which implies that the relationship between the log-odds of class membership
and covariate values is a linear relationship. This assumption may be evaluated
through visual inspection of log-odds plots and through statistical tests like
the Box-Tidwell test. The model also assumes
an absence of multicollinearity within the covariates. This may be tested by
evaluating correlation between covariates and by statistics like the Variance
Inflation Factor (VIF), and will be corrected by dropping highly correlated 
predictors from the selection subset (which may impact analysis conclusions).
This model assumes independence of errors, which may
be evaluated through visual inspection of log-odds plots. However, if this 
assumption is not met (within a reasonable degree), a more drastic change to
model structure will be necessitated to correct for correlation structures within
the data. Lastly, logistic regression assumes that there are no highly influential 
outliers or leverage points, but, given the highly constrained covariate
values innate to survey data, we do not foresee the presence
of such observations.

\qquad The value for each coefficient and the associated log-odds
for each response is typically estimated using maximum likelihood methods. 
The statistical significance for 
coefficient maximum likelihood estimates is typically evaluated using a Wald 
test, where the test statistic is defined as

$$
Z=\frac{\hat{\beta_p}}{s.e.(\hat{\beta_p})}.
$$

The ability to 
specify and easily calculate the likelihood enables the opportunity to use 
information criteria like
the Akaike information criterion (AIC) and Bayesian information criterion (BIC)
to compare models with differing subsets of predictors. As such, each model's
covariates will be selected using forward stepwise selection using BIC as the 
selection criteria. Forward selection was chosen over other selection methods 
(backwards, bidirectional) because some covariates have relatively small sample
sizes for certain classes, so it is preferable to limit the number of covariates
included in the initial models. For example, from Table 2 we see that only 37
individuals who do not describe themselves as white participated in the survey.
Similarly, we will use BIC over AIC to compare between models, as BIC typically
favors more parsimonious models (models with fewer covariates). For this procedure,
an initial model will be chosen, then covariates will be iteratively added to the
model until AIC is no longer substantially reduced.

\qquad For Hypothesis 4, we will use a similar procedure, but logistic regression
will be replaced by linear regression, as we will use the assigned knowledge of
labor rights score as the continuous outcome. Linear regression models are specified as 

$$
Y=\beta_0+\beta_1X_1+\cdots+\beta_pX_p+\epsilon
$$

where $\epsilon$ is distributed $N(0, \sigma^2)$. Here, the assumptions are 
largely similar to those of logistic regression. We assume a linear relationship
between covariates and the outcome, which can be evaluated using visual
inspection of residual plots. Errors are assumed to be independent with constant
variance, which will be evaluated in the same manner. Multicollinearity is
expected to be absent from the covariate set, which may be evaluated and corrected
for in the same manner as logistic regression. And, as with logistic regression,
the presence of influential outliers and leverage points is not expected to
be a problem in this instance.Lastly, as with logistic regression, maximum 
likelihood estimates will be used,
enabling us to use similar test statistics and information criterion for model
selection. So, we will procede with the same forward selection using BIC 
procedure as discussed with logistic regression.

\newpage

## Reproducibility

The code used to generate this analysis is available on GitHub 
at `<Link to be added when/if repository is made public>`
. The `Background` folder contains information pertinent to understanding the
analysis but unnecessary for reproducing it. The `Code` folder contains four
files: `1_cleanData.R`, `2_cleanMultResp.R`, `3_analysisData.R`, and
`4_summaryTables.R`.
These R scripts are dependent upon a data file 
`DataRaw/Cannabis Workers and COVID_Master_3Feb22.xlsx` 
which is not available on GitHub, but may be 
requested by emailing `max` `<dot>` `mcgrath` `<at>` `ucdenver` `<dot>` `edu`. 
To run the complete analysis, 
each script should be run in the order of the number prefixing its filename. The
last directory, `Report`, contains the RMarkdown file `interim.Rmd` which may
be used to generate this report (noting that it also depends on the aforementioned
data and scripts).

The complete details of the `R` version, package versions, and
machine details for the instance which generated this report are provided below.

```{r, echo = FALSE}
sessionInfo()
```


\newpage

```{r}
kable(missingnessSummary, booktabs = TRUE,
      caption = "Missingness Summary", 
      col.names = c("Variable", "Number Missing"),
      align = c('l', 'r')) %>%
    kable_styling(latex_options = c("striped", "scale_down")) %>%
    row_spec(11, hline_after = TRUE)
```

```{r}
kable(outcomeSummary, booktabs = TRUE,
      caption = "Unionization Support Summary by Covariate",
      col.names = c("Category", "Yes", "No", "Yes", "No", "Yes", "No")) %>%
    add_header_above(c(" " = 1, "Safer Workplaces" = 2, "Better Wages" = 2,
                       "Better Benefits" = 2)) %>%
    add_header_above(c(" " = 1, "Support of Unionization for..." = 6)) %>%
    pack_rows("Bio. Hazards", 1, 2) %>%
    pack_rows("Chem. Hazards", 3, 4) %>%
    pack_rows("Phys. Hazards", 5, 6) %>%
    pack_rows("Gender", 7, 8) %>%
    pack_rows("Race", 9, 10) %>%
    pack_rows("Wage", 11, 16) %>%
    pack_rows("Age", 17, 23) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 8) %>%
    column_spec(c(3, 5), border_left = FALSE, border_right = TRUE)
```

```{r}
kable(knowledgeTable, booktabs = TRUE,
      caption = "Knowledge of Workers' Rights Summary by Covariate",
      col.names = c("Category", "Average Score", "Standard Error"),
      digits = 2, align = c("l", "c", "c")) %>%
    pack_rows("Bio. Hazards", 1, 2) %>%
    pack_rows("Chem. Hazards", 3, 4) %>%
    pack_rows("Phys. Hazards", 5, 6) %>%
    pack_rows("Gender", 7, 8) %>%
    pack_rows("Race", 9, 10) %>%
    pack_rows("Wage", 11, 16) %>%
    pack_rows("Age", 17, 23) %>%
    kable_styling(latex_options = c("striped"))
```


```{r}
kable(hazardsTable, booktabs=TRUE, 
      caption = "Workers' Concerns Over Hazards Summary",
      digits = 4, col.names = c("Type of Hazard", "Count", "Percent", "Concerns", 
                                "No Concerns")) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 7) %>% 
    add_header_above(c("Specific Hazards Summary" = 3, "Category Summary" = 2)) %>%
    column_spec (4, border_left = TRUE, border_right = FALSE) %>%
    pack_rows("Biological Hazards", 1, 4) %>%
    pack_rows("Chemical Hazards", 5, 15) %>%
    pack_rows("Physical Hazards", 16, 37)
```

\newpage

