---
title: "A Statistical Analysis of Occupational Health Concerns and Workers' Rights in the Cannabis Space"
author: |
    | Analyst: Max McGrath
    | Investigator: Dr. Marty Otañez
    | MATH 6330 - Workshop in Statistical Consulting
    | Report generated: `r format(Sys.Date(), '%m/%d/%Y')`
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{titling}
    - \usepackage{fixmath}
    - \usepackage{amsmath}
output: 
    pdf_document:
        number_sections: true
indent: false
mainfont: Times New Roman
fontsize: 12pt
geometry: margin=1in
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')
library(kableExtra)
setwd("..")
source("Code/4_summaryTables.R")
source("Code/5_modelSupport.R")
source("Code/6_modelKnowledge.R")
```

# Introduction

We received responses to 102 survey questions from 213 individuals who currently
work in the Colorado cannabis industry or who were employed in the Colorado 
cannabis industry in the past 12 months. Most relevant to this analysis, 
respondents were asked whether they support unionization within the cannabis 
industry for three purposes: creation of safer workplaces, negotiation of better
wages, and negotiation of better benefits. Further, each respondent's ability to
recognize existing federal labor protections was assessed, and they were asked 
whether they have concerns regarding their exposure to three types of workplace 
hazards: biological, chemical, and physical. Lastly, they were asked their age, 
gender, race, and hourly compensation.

\qquad This study aims to investigate the relationships between cannabis 
employees' support and knowledge of labor rights and their demographics, wages, 
and concerns over exposure to occupational hazards by testing the following four
hypotheses:

1. Cannabis workers who express concern over exposure to occupational hazards are
more likely to express support for labor rights.
2. Female cannabis workers are more likely to support workers’ rights than their
male counterparts.
3. Non-white cannabis workers are more likely to support workers’ rights than 
their white counterparts.
4. Higher paid cannabis employees show more knowledge about workers’ rights than
lower paid employees.

# Analysis and Results

Of the 205 individuals included in this analysis, approximately
75.1% support unionization for safer workplaces, 82.0% support unionization for
better wages, and 82.4% support unionization for better benefits. Further, 56.1%
expressed concern over some type of biological hazard, 60.0% expressed concern
for some type of chemical hazard, and 68.8% expressed concern for some type of
physical hazard present in their workplace. Approximately 58% of respondents
were female or non-binary and 82% are white. The two most represented wage
categories are those who make $14 dollars or less per hour with 30.7% of respondents 
and those whose make $19 or more per hour with 24.9% of respondents. The most
represented age category is 26 to 30 years old with 36.6% of respondents. A complete summary
of the number of respondents within each category stratified by support for 
unionization is provided in Table 1. Each respondent was assigned a knowledge of
workers' rights scores, with details provided in Section 4.1. The average knowledge
score was 1.16 correctly answers. Table 2 provides the 
mean and standard error of knowledge scores stratified by covariate values.

```{r}
kable(outcomeSummary, booktabs = TRUE,
      caption = "Unionization Support Summary by Covariate",
      col.names = c("Category", "Yes", "No", "Yes", "No", "Yes", "No")) %>%
    add_header_above(c(" " = 1, "Safer Workplaces" = 2, "Better Wages" = 2,
                       "Better Benefits" = 2)) %>%
    add_header_above(c(" " = 1, "Support of Unionization for..." = 6)) %>%
    pack_rows("Bio. Hazards", 1, 2) %>%
    pack_rows("Chem. Hazards", 3, 4) %>%
    pack_rows("Phys. Hazards", 5, 6) %>%
    pack_rows("Gender", 7, 8) %>%
    pack_rows("Race", 9, 10) %>%
    pack_rows("Wage", 11, 14) %>%
    pack_rows("Age", 15, 19) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 8) %>%
    column_spec(c(3, 5), border_left = FALSE, border_right = TRUE)
```


```{r}
kable(knowledgeTable, booktabs = TRUE,
      caption = "Knowledge of Workers' Rights Summary by Covariate",
      col.names = c("Category", "Average Score", "Standard Error"),
      digits = 2, align = c("l", "c", "c")) %>%
    pack_rows("Bio. Hazards", 1, 2) %>%
    pack_rows("Chem. Hazards", 3, 4) %>%
    pack_rows("Phys. Hazards", 5, 6) %>%
    pack_rows("Gender", 7, 8) %>%
    pack_rows("Race", 9, 10) %>%
    pack_rows("Wage", 11, 14) %>%
    pack_rows("Age", 15, 19) %>%
    kable_styling(latex_options = c("striped"))
```


\qquad To evaluate Hypotheses 1-3, three logistic regression models were fit and 
refined, with each modeling one of the three types of unionism support. The 
final coefficient estimates for each model are provided in Table 3. From these 
results, we see that an individual expressing concern over exposure to physical 
hazards was positively associated with their odds of supporting unionism for all
three reasons: safer workplaces, better wages, and better benefits. For example,
based on the "Safer Workplaces" model, the odds of an individual supporting 
unionization within the cannabis space for the purpose of creating safer 
workplaces is 3.82 times greater if they expressed concerned over physical 
hazards compared with someone who did not (p-value < 0.001). We also see 
significant evidence that non-white individuals have 8.22 times greater odds, on
average, of supporting unionization for the purposes of negotiating better wages
when controlling for concerns over physical and biological hazards 
(p-value = 0.045). Additionally, there is weaker evidence that non-white 
individuals also have greater odds of supporting unionization for safer 
workplaces and better benefits when controlling for concerns over physical
hazards (p-values 0.064 and 0.076, respectively). Throughout the model selection
process, gender was not found to be significantly associated with any type of 
support for unionization in the cannabis industry.

\newpage

\qquad To evaluate Hypothesis 4, a linear regression model was fit and refined
using assigned knowledge of workers' rights score as the outcome.
The resulting coefficient estimates are provided in Table 4. From these results,
we see that individuals who make \$19 or more an hour had, on average, a 0.578 
higher knowledge of workers' rights score relative to individuals who make 
$14 or less per hour when controlling for gender, age, and race 
(p-value = 0.014). We also see that non-white individuals had a 0.748 higher
score than white individuals (p-value = 0.001), people aged 31 to 35 and 36 to 
40 score higher than those under 25 (p-values = 0.005 and 0.023, respectively), 
and individuals who identify as female or non-binary had a 0.677 lower score 
than those who identify as male when controlling for other covariates 
(p-value < 0.001). 

```{r}
kable(unionSupportSummary[, 2:5], booktabs = TRUE,
      caption = "Support for Unionization Logistic Regressions Coefficient Summary by Model",
      col.names = c("Outcome/Variable", "Coef. Estimate", "95% CI", "p-value")) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 8) %>%
    pack_rows("Safer Workplaces", 1, 2) %>%
    pack_rows("Better Wages", 3, 5) %>%
    pack_rows("Better Benefits", 6, 7)
```

```{r}
kable(knowledgeSummaryData[, 2:5], booktabs = TRUE,
      caption = "Knowledge of Workers' Rights Linear Regression Coefficient Summary",
      col.names = c("Variable", "Coef. Estimate", "95% CI", "p-value")) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 8) %>%
    pack_rows("Gender", 1, 1) %>%
    pack_rows("Age", 2, 5) %>%
    pack_rows("Race", 6, 6) %>%
    pack_rows("Wage", 7, 9)
```

# Discussion

From the above results, we see evidence supporting Hypothesis 1, as concern over
exposure to physical hazards was positively associated with all types of 
unionism support. There is also weaker evidence that concern over biological 
hazards is associated with support for unionization for the purpose of 
negotiating better wages when controlling for exposure to physical hazards. 

\qquad It is noteworthy that all three types of workplace hazard concerns were 
included as potential predictors in each of the three models. So, when the model
selection process chose concern over physical hazards as the best predictor of
unionism support, it means only that other types of hazard concerns
no longer improved the model once concern for physical hazards had been 
accounted for. Additional exploratory models were fit in which only one type of
hazard concern was included at a time, and, for each outcome, all three types of
hazard concerns were retained in the final models with a positive coefficient. 
As such, there is an indication that each type of workplace hazard concern is 
positively associated with each type of unionism support.

\qquad It is also notable that the physical hazards category contained the 
greatest number of unique survey response options of the three types of hazard
concerns (see Table 5). And while physical hazards had the greatest overall
number of respondents express concern, the specific sub-categories of hazards in the
physical category generally saw fewer responses than those in other 
categories. The sub-categories of hazards in the physical category
had an average of 36.05 respondents express concern, while those in the chemical
and biological categories had averages of 43.7 and 64, respectively.
As such, the quantity of sub-categories of hazards contained within the physical 
hazards category may be driving its statistical importance over the 
other two categories. As such, refining the physical category into multiple 
smaller categories or reducing its scope in future surveys and analyses may yield
more clear insights into the specific types of hazard concerns that are most
associated with support on unionization within the cannabis industry. 

```{r}
kable(hazardsTable, booktabs=TRUE, 
      caption = "Workers' Concerns Over Hazards Summary",
      digits = 4, col.names = c("Type of Hazard", "Count", "Percent", "Concerns", 
                                "No Concerns")) %>%
    kable_styling(latex_options = c("striped", "scale_down"),
                  font_size = 7) %>% 
    add_header_above(c("Specific Hazards Summary" = 3, "Category Summary" = 2)) %>%
    column_spec (4, border_left = TRUE, border_right = FALSE) %>%
    pack_rows("Biological Hazards", 1, 4) %>%
    pack_rows("Chemical Hazards", 5, 15) %>%
    pack_rows("Physical Hazards", 16, 37)
```

\qquad The results do not support Hypothesis 2, as there is no indication that gender
is associated with any type of unionization support. However, these results do
lend credence to Hypothesis 3. We found significant evidence that non-white 
individuals are associated with greater odds of supporting unionization to negotiate 
better wages relative to white individuals when holding physical hazard
concerns constant. Similarly, we see weaker evidence that non-white individuals
are associated with greater odds of supporting unionization for the purposes of
creating safer workplaces and negotiating better wages. Given the relatively low
sample size for non-white individuals (18.0%), increasing the number of non-white 
individuals in future surveys may better elucidate the relationship between
race and unionization support.

\qquad Finally, the results of our final model support Hypothesis 4, as we see that
individuals who make $19 or more are associated with an increase in the
number of federal protections they were able to identify relative to those
who make $14 or less, noting that neither the \$15-16 or \$17-18 groups performed
significantly better or worse than the \$14 or less group. 
We also see additional evidence that individuals aged
31-40 are associated with greater scores relative to those under 25, non-white 
individuals are associated with higher scores relative to white individuals, and
males are associated with higher scores relative to those who identify as female
or non-binary.

\qquad One noteworthy aspect of the analysis of knowledge scores stems from the
inclusion of the survey answer "Don't know". Given that the average score across
all subjects was only 1.16 on a scale from 0 to 4, we expect a person guessing 
would have had an above
average score. So, if certain groups guessed more than others, it may 
have skewed analysis results. To briefly investigate this, we looked at the 
average number of incorrect answers for each group, as we would expect to see
inflated numbers of incorrect answers if any group had guessed more than
others rather that truthfully selecting "Don't know". However, in each case, 
groups which are significantly associated with higher scores in the above 
results also had  lower average numbers of incorrect scores than the group 
against which they were compared. For example, non-white individuals had an 
average of 0.405 incorrect scores while white individuals had an average of 0.5 
incorrect answers. While this limited analysis suggests that our results
were not impacted by differential levels of guessing, it may be 
worthwhile to consider the potential for guesswork impacting future assessments 
of respondents' knowledge of workers' rights and to update survey questions
accordingly.

# Methods

## Data Preparation

First, incomplete observations were removed from the data set. These incomplete
observations came from the questions about gender and race, as responses
where individuals chose "prefer not to say" for either question 
were marked as missing. Gender had 4 missing values and Race had 7 missing
values (with some overlap), so in total 8 observations were removed for a final
sample size of 205. All other relevant questions had no missing 
responses. 

\qquad It is also notable that the question asking survey takers' gender identity
included a 
"non-binary/third gender" option. So, for the purposes of this analysis, the gender 
category was reduced to a binary category in which respondents were classified 
as those who identify as male and those who identify as female, 
non-binary, or third gender. Similarly, the question 
that asked respondents to describe their race
included seven options, and that category was reduced to a binary
category classifying respondents as those who describe themselves as Asian, 
Black, Native American, Native Alaskan, Native Hawaiian, or Pacific Islander 
and those who describe themselves as white. Age
and hourly compensation (wage) are both categorical variables with 6 and 7
possible responses, respectively. Lists of possible responses are included as a
part of Tables 1 and 2, noting that the age category offered several responses
for ages above 41 that were reduced to a single "41+" category due to
limited responses in the those categories. Similarly, the wage categories 
"\$11-12" and "\$13-14" were combined to create "$14 or less", and "\$19-20"
and "\$21 or more" were combined to create "$19 or more".

\qquad Measurements of workers' concerns over occupational hazards stem from 
their answers to the following three multiple response questions: 

* What __biological__ hazards exist in your cannabis workplace that concern
you?
* What __chemical__ hazards exist in your cannabis workplace that concern you?
* What __physical__ hazards exist in your cannabis workplace that concern you? 

Each of these questions allowed users to select multiple responses from a list
of relevant types of hazards contained within each category. For the purposes of
this analysis, each respondent
was given a binary assignment for each category of hazard which indicates 
whether they expressed concern for any hazard in that category. A complete
description of each category of hazard is provided in Table 5. 

\qquad Cannabis workers' support for unionism was measured based on their 
responses to the three questions:

* Do you believe that cannabis workers need to be part of a labor union
union to __create safer workplaces__?
* Do you believe that cannabis workers need to be part of a labor union
union to __negotiate better wages__?
* Do you believe that cannabis workers need to be part of a labor union
union to __negotiate better benefits__?

Each of these questions had two possible responses, "yes" and "no". Each
question's set of responses will be analyzed separately so that
differing motivations for support of unionism may be elucidated by this
analysis. A complete description of respondents' answers to these questions is
provided in Table 1, with counts stratified by values for each the above
covariates.

\qquad Lastly, respondents' knowledge of workers' rights was assessed
by their ability to recognize four existing federal protections. The
individuals surveyed were asked:

* Does federal law protect workers under the worker protection standard 
from retaliation from bringing forward safety concerns?
* Does federal law allow cannabis workers to form a labor union?
* Does federal labor law prohibit cannabis employers from taking any 
adverse employment action against you for engaging in concerted activity
regarding wages, benefits and other terms and conditions of 
employment?
* Do federal labor laws protect you in engaging in concerted activity even 
if you are not in a labor union?

Each question had response options of "yes", "no", and "I don't know". The
respondents were provided an example of concerted activity which says, 
"complaining about safety, having discussion with co-workers about working 
conditions."
Noting that the correct response to each question is "yes", each respondent was
assigned a score from 0 to 4, with one point being awarded for each correct
response. Table 2 provides the sample mean (and standard error) for assigned
scores stratified by each covariate of interest.

## Analysis

First, summary statistics for covariates and outcomes were calculated. Then,
to evaluate Hypotheses 1-3, three logistic regression models, one for each type
of unionism support, underwent forward stepwise model selection. Then, each 
model's final coefficients and their
corresponding estimates and statistical significance were used to determine 
whether the data falsifies or lends evidence to each of the above hypotheses.
For Hypothesis 4, we used a similar procedure, but logistic regression
was replaced by linear regression, as we used respondents' assigned knowledge of
labor rights score as the continuous outcome. As before, coefficient estimates
and their associated statistical significance were used to evaluate the 
hypothesis. An significance threshold of $\alpha$ = 0.05 was used, but, given 
that this is an exploratory analysis, we refer to results that have an associated
p-value near 0.05 as "weaker evidence." For more details on logistic and linear 
regression, see Witten et al. (2013). 

\qquad All pre-processing and analysis was performed using R version 4.1.2 
and RStudio (R Core Team, 2021). Pre-processing was performed using `dplyr`
version 1.0.8 (Wickham et al., 2022) and `tidyr` version 1.2.0 
(Wickham and Girlich, 2022). Modelling was performed using using the base
package `stats`, and forward selection was performed using `MASS` version 
7.3.54. Complete details of the instance used to generate this analysis and 
report are provided in Appendix Section 6.2.

\newpage

# References

| James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013). _An introduction to statistical_
|       _learning_ (1st ed.) [PDF]. Springer.

| R Core Team (2021). _R: A language and environment for statistical computing_. R Foundation 
|       for Statistical Computing, Vienna, Austria. https://www.R-project.org/.

| Venables WN, Ripley BD (2002). _Modern Applied Statistics with S_, Fourth edition. Springer, 
|       New York. ISBN 0-387-95457-0, https://www.stats.ox.ac.uk/pub/MASS4/.

| Wickham H, François R, Henry L, Müller K (2022). _dplyr: A Grammar of Data Manipulation._ 
|        https://dplyr.tidyverse.org, https://github.com/tidyverse/dplyr

| Wickham H, Girlich M (2022). _tidyr: Tidy Messy Data_ https://tidyr.tidyverse.org, 
|        https://github.com/tidyverse/tidyr.

\newpage

# Appendix

## Methods

First, summary statistics for covariates and outcome were calculated. The majority,
of these statistics were simple counts or means, but we did calculate the 
standard error of the mean knowledge score as

$$
\frac{s}{\sqrt{n}}
$$

where $s$ is the sample standard deviation of the knowledge score for each group
and $\sqrt{n}$ is the sample size of each group.

To evaluate Hypotheses 1-3, three logistic regression models, one for each type
of unionism support as the outcome,
underwent model selection. All three types of hazard exposure, gender, race, age
and wage were included as potential predictors for each model. 
Then, each model's final coefficient estimates and statistical significance was 
used to determine 
whether the data disputes or lends evidence to each of the above hypotheses.

\qquad A logistic regression model is specified as

$$
P(X)=\frac{e^{\beta_0+\beta_1X_1+\beta_2X_2+\cdots+\beta_pX_p}}
{1+e^{\beta_0+\beta_1X_1+\beta_2X_2+\cdots+\beta_pX_p}}
$$

where $P(X)=P(Y=1|X)$ represents the probability that an individual belongs to 
a certain outcome class given their observed covariate values. In this analysis,
we will use binary outcomes ("yes" = 1, "no" = 0). 
The covariates
in a logistic regression model may be continuous, ordinal, or binary, but we 
will only
be using binary and ordinal covariates. The above
equation is typically rewritten as

$$
\log\left(\frac{P(X)}{1-P(X)}\right)=\beta_0+\beta_1X_1+\cdots+\beta_pX_p
$$

which implies that the relationship between the log-odds of class membership
and covariate values is a linear relationship. This assumption was evaluated
through visual inspection of log-odds plots, noting that it is unlikely this
assumption would not be met given that only binary variables remained in the
final models. The model also assumes
an absence of multicollinearity within the covariates. This was tested by
looking at the Variance Inflation Factor (VIF). All VIF's were close to one,
so no change was necessitated.
This model assumes independence of observations, however this assumption is 
difficult to test outside of known dependence structures (e.g. time-series or
spatial dependence). But, in the absence of such structures, as is the case with
our data, we are not concerned that this assumption is violated. Lastly, logistic 
regression assumes that there are no highly influential 
outliers or leverage points, but, given the highly constrained covariate
values innate to our survey data, this is unlikely to be a problem.

\qquad The value for each coefficient and the associated log-odds
for each response is typically estimated using maximum likelihood methods. 
The statistical significance for 
coefficient maximum likelihood estimates is typically evaluated using a Wald 
test, where the test statistic is defined as

$$
Z_p=\frac{\hat{\beta_p}}{s.e.(\hat{\beta_p})}.
$$
We then calculate 95% confidence intervals as

$$
CI=\hat{\beta}_p\pm1.96*s.e.(\hat{\beta}_p).
$$

The ability to 
specify and easily calculate the likelihood enables the opportunity to use 
information criteria like
the Akaike information criterion (AIC) and Bayesian information criterion (BIC)
to compare models with differing subsets of predictors. As such, each model's
covariates were selected using forward stepwise selection using BIC as the 
selection criteria. Forward selection was chosen over other selection methods 
(backwards, bidirectional) because some covariates have relatively small sample
sizes for certain classes, so it is preferable to limit the number of covariates
included in the initial models. For example, from Table 1 we see that only 37
individuals who do not describe themselves as white participated in the survey.
Similarly, we will use BIC over AIC to compare between models, as BIC typically
favors more parsimonious models (models with fewer covariates). For this procedure,
an initial model was chosen using the single predictor which reduced the 
BIC the most over the intercept only model. Then, covariates were iteratively
added to the model based on which one reduced the BIC the most over the previous
model until BIC was no longer reduced by including additional predictors. All
three types of exposure to hazards, gender, race, age, and wage were
considered as potential predictors in this process.

\qquad For Hypothesis 4, we used a similar procedure, but logistic regression
was replaced by linear regression, as we used the assigned knowledge of
labor rights score as the continuous outcome. Linear regression models are specified as 

$$
Y=\beta_0+\beta_1X_1+\cdots+\beta_pX_p+\epsilon
$$

where $\epsilon$ is distributed $N(0, \sigma^2)$. Here, the assumptions are 
largely similar to those of logistic regression. We assume a linear relationship
between covariates and the outcome, which was evaluated using visual
inspection of residual and QQ plots. Errors are assumed to be independent with constant
variance, which was evaluated in the same manner. Multicollinearity is
expected to be absent from the covariate set, which was evaluated and corrected
in the same manner as logistic regression. And, as with logistic regression,
the presence of influential outliers and leverage points is not expected to
be a problem in this instance. Similarly to logistic regression, inference will
be conducted using analogous Wald statistics and confidence intervals. 
Lastly, as with logistic regression, maximum 
likelihood estimates will be used,
enabling us to use similar test statistics and information criterion for model
selection. So, we will proceed with the same forward selection using BIC 
procedure as discussed with logistic regression. Here, only gender, race, wage,
and age were considered as potential predictors.

\qquad We will note here that our outcome for the linear model was not truly 
continuous, taking discrete values from 0 to 4. Further, our predictor set
was composed entirely of categorical and dichotomous variables. This led to
"blocky" behavior in the residuals which was cause for some concern. To ensure
the robustness of our results, we repeated the analysis 5 times, each time
removing 10 random observations from the data. In each case, the results
remained consistent, lending credibility to our results despite imperfect 
residual structure.

\newpage

## Reproducibility

The code used to generate this analysis is available on GitHub 
at `<Link to be added when repository is made public>`
. The `Background` folder contains information pertinent to understanding the
analysis but unnecessary for reproducing it. The `Code` folder contains four
files: `1_cleanData.R`, `2_cleanMultResp.R`, `3_analysisData.R`,
`4_summaryTables.R`, `5_modelSupport.R`, `6_modelKnowledge.R`, and 
`7_additionalAnalysis.R`.
These R scripts are dependent upon a data file 
`DataRaw/Cannabis Workers and COVID_Master_3Feb22.xlsx` 
which is not available on GitHub, but may be 
requested by emailing `max` `<dot>` `mcgrath` `<at>` `ucdenver` `<dot>` `edu`. 
To run the complete analysis, 
each script should be run in the order of the number prefixing its file name. The
last directory, `Report`, contains the RMarkdown file `report.Rmd` which may
be used to generate this report (noting that it also depends on the aforementioned
data and scripts).

The complete details of the `R` version, package versions, and
machine details for the instance which generated this report are provided below.

```{r, echo = FALSE}
sessionInfo()
```



